---
title: "Average two or more columns from a data frame"
author: "Francesco Bailo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(knitr)
require(tidyverse)
```

Let's say that I have some survey data, where each row defines the responses from a single respondent. This is of course a very common way to share survey data. 

Let's also say that I have two or more continuous variables in the data **that I want to average** to create a new composite variable.

## Get some data

We first create a data frame with some random data simulating four survey responses on a value from 1 to 5 from 10 different *respondents*.

```{r}
survey_data <- data.frame(
  respondent_id = 1:10,  # 10 respondents
  question_1 = sample(1:5, 10, replace = TRUE),
  question_2 = sample(1:5, 10, replace = TRUE),
  question_3 = sample(1:5, 10, replace = TRUE),
  question_4 = sample(1:5, 10, replace = TRUE)
)
```

In case you are interested, let's break down the functions we are using here:

- `data.frame()` create a data frame object in R, where each attribute `name = c(..values..)` is going to be a column. 
- `1:10` Simply create a vector of numbers from 1 to 10. You can of course, use different combinations (e.g, `-1:42`)
- `sample()` with randomly sample from a vector. In this case, we pass three attributes:
  1. `1:5` define the vector (or population) from which to choose. In this case is the vector `c(1, 2, 3, 4, 5)` because we want to simulate a common situation where survey answers are given using a [Likert-type scale](https://en.wikipedia.org/wiki/Likert_scale) (e.g., 1 = Strongly disagree; 2 = Disagree; 3 = Neither agree nor disagree; 4 = Agree; 5 = Strongly agree).
  2. `10` defined how many "individuals" to sample from our "population". In this case we decided we wanted 10 respondents and so we we will need to set a response to each question for each respondent.
  3. `replace = TRUE` means that after we sample the first value (either, 1, 2, 3, 4 or 5) we put the value back into the population so that can be re-sampled again. This is sometimes called "sampling with replacement". 

## Average columns

Now that we have some data, we can average existing variables (i.e. columns) into new variables (i.e. columns).

Let's say I want to create a new variable as the average responses of `question_2` and `question_4`:

```{r}
survey_data$new_variable_1 <- # Create new variable and `<-` assign some value
  (survey_data$question_2 + survey_data$question_4) / 2 # Sum columns and divide
```

Or let's say I want to average the responses from question 1, 3, and 4:

```{r}
survey_data$new_variable_2 <-
  (survey_data$question_1 + survey_data$question_3 + survey_data$question_4) / 3
```

(Note that this time we divide by 3 since we have three variables)

Let's now see the results (note that if you replicate this you will have different numbers has `sample()` produces each time a *random* sample!)

```{r eval = FALSE}
head(survey_data)
```

```{r echo = FALSE}
survey_data %>%
  kable()
```



## Issue 1: What if we have missing values?

If you have missing value in your responses, some of your composite value are also going to be missing (NA).

For example let's say that the first respondent didn't respond to question 1. 

We can replace the current value of the answer with a `NA` to reflect this

```{r}
survey_data$question_1[1] <- NA
```

```{r echo = FALSE}
survey_data %>%
  kable()
```

## Issue 2: What if the scales of the responses are different?

So far we have considered responses with the same scale (1-to-5). Let's say one of the questions implied a different scale (e.g, 1-to-10). Let's consider a different data frame

```{r}
survey_data <- data.frame(
  respondent_id = 1:100,  # 100 respondents
  question_1 = sample(1:5, 100, replace = TRUE), # Note the 1-to-5 scale
  question_2 = sample(1:10, 100, replace = TRUE) # Note the 1-to-10 scale
)
```

Can we still average them all together? Well, no. That is probably not appropriate as the wider scale is going to average out the scale with the narrower scale. 

- We need to rescale! How do we do it?

The easiest is probably to rescale everything to a common scale - 0 to 1 (which of course can be easily converted to a 0 to 100 scale multiplying by 100...) -  before averaging.

In order to rescale a value from a known scale into a 0-to-1 scale, we can do this

```{r eval = F}
min_val <- 1 # Minimum value of the original scale
max_val <- 5 # Maximum value of the original scale
old_val <- 4 # Let's say for example the value to rescale is `4`

# Formula   
(old_val - min_val) / (max_val - min_val)
```

Let's rescale both variables

```{r}
min_val <- 1 # Minimum value of the original scale
max_val <- 5 # Maximum value of the original scale
survey_data$question_1_rescaled <- 
  (survey_data$question_1 - min_val) / (max_val - min_val)
```

and

```{r}
min_val <- 1 # Minimum value of the original scale
max_val <- 10 # Maximum value of the original scale
survey_data$question_2_rescaled <- 
  (survey_data$question_2 - min_val) / (max_val - min_val)
```

Let's compare the distributions to see if everything is in order!

```{r}
ggplot(survey_data, aes(x = question_1)) +
  geom_histogram()
```

which should have the exact same distribution of

```{r}
ggplot(survey_data, aes(x = question_1_rescaled)) +
  geom_histogram()
```
