---
title: "Join data"
author: "Francesco Bailo"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
---

Let's assume I have data in two different data frames but I wan them in the same data frame. As this is a very common problem, there are many ways to do this (just have a look at the number of highly voted answers to [this question](https://stackoverflow.com/questions/1299871/how-to-join-merge-data-frames-inner-outer-left-right)).

What you want is to **join** two data frames.

## Option 1: The dplyr solution 

dplyr using the [SQL](https://en.wikipedia.org/wiki/SQL) vocabulary, offers four *join* verbs. Which one you need depends on what you want to achieve.

* `dplyr::left_join(x, y)` keeps all observations in table x.

<img src = 'https://www.w3schools.com/sql/img_leftjoin.gif'></img>

* `dplyr::right_join(x, y)` keeps all observations in table y.

<img src = 'https://www.w3schools.com/sql/img_rightjoin.gif'></img>

* `dplyr::full_join(x, y)` keeps all observations in x and y.

<img src = 'https://www.w3schools.com/sql/img_fulljoin.gif'></img>

* `dplyr::inner_join(x, y)` only keeps obervations in x with a matching key in y. 

<img src = 'https://www.w3schools.com/sql/img_innerjoin.gif'></img>

Most likely, if you want to enrich an existing table with some additional data what you are looking for is a `left_join()`.

```{r message = F, warning=F}
library(dplyr)
```

Let's consider these two tables

```{r}
band_members <- 
  data.frame(member = c("Mick", "John", "Paul"),
             band = c("Stones", "Beatles", "Beatles"))
band_members

band_instruments <- 
  data.frame(player = c("John", "Paul", "Keith"),
             instrument = c("guitar", "bass", "guitar"))
band_instruments
  
```

I can join them with using the column `name`

```{r}
joined_table <-
  dplyr::left_join(band_members, band_instruments,
                   by = c(member = 'player')) #<<
```

The argument `by = c(member = 'player')` identifies the two *index* columns for the matching. `member` (without quotes) identifies the index column in the first table while `'player'` (with quotes) identifies the corresponding index in the second column. 

Let's check the result

```{r}
joined_table
```


Note that,

1. `Mick` from the *first* table is still present in the resulting table but has no instrument (`NA`) since he doesn't appear in the *second* table;

2. `Keith` from the *second* table is not present in the resulting table since it only appears in the *second* table (and is not present in the *first* table).

## Option 2: The base solution 

A similar result can be obtained without using dplyr. You can instead use the `match()` function.

```{r}
band_members$instrument <- 
  band_instruments$instrument[match(band_members$member,
                                    band_instruments$player)]
```

`match(x, y)` will return the positions of the values found in vector `x` in vector `y`. 


