---
title: "Reverse a scale"
author: "Francesco Bailo"
date: "`r Sys.Date()`"
output: html_document
---

## Prepare data

`var_a` is a numeric variable with range 1 to 4:

```{r}

my_dat <- data.frame(var_a = c(1, 1, 1, 2, 2, 4, 1, 3, 2, 4, 1, 2, 1))

```

```{r}

library(ggplot2)

ggplot(my_dat, aes(x = var_a)) +
  geom_bar()

```

<mark>Task: Reverse the scale so that 4 is now 1 and 1 is now 4 (and 2 is now 3 and 3 is now 2).</mark> 

## First approach: Replacing values one by one in a new variable

### Step 1: Create a new variable with the same values as the original one

```{r}

my_dat$var_b <- my_dat$var_a

```

### Step 2: Replace all the values 

(note that the variable within `[...]` is the old one!)

```{r}

my_dat$var_b[my_dat$var_a == 1] <- 4
my_dat$var_b[my_dat$var_a == 2] <- 3
my_dat$var_b[my_dat$var_a == 3] <- 2
my_dat$var_b[my_dat$var_a == 4] <- 1

```

### Done!

```{r}

ggplot(my_dat, aes(x = var_b)) +
  geom_bar()

```

## Second approach: The fancy way

### Step 1: Subtract the max value that the variable can take + 1 to the same variable

```{r}

my_dat$var_c <- max(my_dat$var_a) + 1 - my_dat$var_a

```

Let's check, (never trust your teacher!)

```{r}

my_dat$var_c == my_dat$var_b

```


```{r}

ggplot(my_dat, aes(x = var_c)) +
  geom_bar()

```


Here why this works:

* `5-4=1`
* `5-3=2`
* `5-2=3`
* `5-1=4`
