---
title: "Regression tables with Stargazer"
author: "Francesco Bailo"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

You can download the source RMD to produce PDF and HTML files here

* [stargazer.Rmd](stargazer.Rmd)



## Installing Stargazer 

Make sure to install the package first!

```{r eval = FALSE}
install.packages("stargazer")
```

The stargazer package helps you to print out nicely edited tables from your regressions.

## Step 1. Let's produce some results from a linear model

```{r}
fit <- lm(iris$Sepal.Length ~ iris$Sepal.Width + iris$Petal.Width)
```

You can get a summary of the regression statistics with

```{r}
summary(fit)
```

## Step 2a. Let's present the results in a nice table (in a PDF)

Let's load the package first...

```{r}
library(stargazer)
```

... now we can use the function `stargazer()` with our `lm` object named `fit`.

(You won't see any output if you are not reading this from a PDF)

```{r results = 'asis'}
stargazer(fit, header = FALSE, type='latex')
```


### Notes

We need to specify two options/arguments for this to run properly.

1. We need to specify the option `results = 'asis'` for our code chunk: `{r results = 'asis'}`. This writes "the raw text results directly into the output document without any markups" (see [here](https://yihui.org/knitr/options/#text-output) for details).

2. We need to set the argument `header = FALSE` in the `stargazer()` function, so to avoid adding the name and version of the package just before the table.

3. We can specify the type of output with want. In this case, we need a `type='latex'` as the engine to produce the PDF will need a \LaTeX  formatted document. 

## Step 2b. Let's present the results in a nice table (in a HTML document)

(The output looks good only if you are reading this in a browser)

```{r results = 'asis'}
stargazer(fit, header = FALSE, type='html')
```

1. Just change to `type='html'` and then Knit an HTML file (in RStudio "Knit" and then "Knit to HTML"). 

## Step 2c. Let's present the results in a nice table (for a Word document)

Unfortunately you can't directly produce the table in a Word document created using the "Knit" command. But there is a workaround.

1. Create the table in an HTML document (previous step).

2. Open the HTML in your browser.

3. Copy the table and paste it into a Word document (or a PowerPoint). 



