---
title: "Data visualisation and communication exercise"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document: default
  
---

## General task 

Use R and survey data from 2019 Round 7 of the [Afrobarometer](https://www.afrobarometer.org/data/merged-data/) (34 countries) to study

a. the support for democracy and 

b. confidence in government institutions across Africa. 

### Subtasks

Create two indices:

1. A *support for democracy index* named in your code `support_demo_index`, and 

2. a *confidence in government institutions index* named in your code `confidence_govt_index`.

For each index, 

* Identify, list and discuss the variables that made up the index. 

* (If necessary) recode the variable before creating your indices.

* Construct two different versions of the indices by 

    * averaging your variables (in your code `support_demo_index.avg` and `confidence_govt_index.avg`); and
    
    * applying PCA to the same variables (in your code `support_demo_index.pca` and `confidence_govt_index.pca`).

## Report

1. Describe  the source variables  used  for  making  your indices  using  appropriate  descriptive  statistics  (min, mean, median, st.deviation, max, number of observations) and graphs

2. Briefly  discuss  the  ideas  behind  the  two  approaches  (averaging  and  PCA)  to  index  construction. What do they have in common? How are they different?

3. Discuss the types of the variables you are using in your indices (e.g. binary, categorical, continuous) and discuss any transformation you applied to the variables while constructing your indices.

4. Describe (using visuals and words) the distribution of 

    4.1 `support_demo_index` and 
    
    4.2 `confidence_govt_index`  –  for selected (3+) countries  separately.  
    
    Do  they  look  the  same  across countries? 

5. Conclude with a general comment on support for democracy and confidence in institutions in these countries. 

## R programming tips

### Reading an SPSS data file

Survey responses are stored in an SPSS data files (.sav). To import then in your R session you can do

```{r}
library(haven)
r7_merged_data_34ctry_release <- 
  # Change path accordingly
  read_sav("r7_merged_data_34ctry.release.sav") 
```

### Labelled variables

Most of the variables are `double` (numerical) and `labelled`.

```{r}
class(r7_merged_data_34ctry_release$Q43D)
```

Before using any variable, you need to check the mapping between numerical values and labels. Use `haven::print_labels()` on any variable to check the `value`-`label` mapping.

```{r}
haven::print_labels(r7_merged_data_34ctry_release$Q43D)
```

### Recoding labelled variables

If you want to **recode a labelled numerical variable into a factor variable** where numeric values are replaced by character values, you can use `haven::as_factor()`.

So,

```{r}
r7_merged_data_34ctry_release$Q43D[1:10]
```

becomes...

```{r}
haven::as_factor(r7_merged_data_34ctry_release$Q43D[1:10])
```

If you want to **use a labelled numerical variable as a quantitative discrete variable** to get some descriptive statistics (e.g. the mean), you might still need to recode it (e.g. with `dplyr::recode()`) - again do `haven::print_labels()` on the variable - as likley not every response should be included in your quantitative analysis. 

## Submission guidelines

Produce a PDF report based on your analysis using the R markdown document (.Rmd) template provided. 

For this assignment your are expected to submit:

1. **Your report** Render your report into a PDF from the source .Rmd document. Make sure to `echo = TRUE` all your chunks in final the PDF, so that your code is directly visible in your PDF.

2. **Your R project folder** Compress your project folder, which is expected to include:

    2.1. **Your .Rmd document**;
  
    2.2. **The data file**;
  
    2.3. **Your .Rproj file**.
    
**Length**: 1500wd (as measured in the R Markdown document)