---
title: "[SSPS4102 A1 template: Replace with a meaningful title]"
author: "[STUDENT NAME: Replace with your name]"
output:
  bookdown::pdf_document2: default
date: "`r Sys.Date()`"
---

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

```

<!-- Delete from this line until the end of the document before submitting --> 

## How do I create and reference my figures?

See Figure \@ref(fig:cars-plot).

```{r cars-plot, fig.cap="This is your caption"}
par(mar = c(4, 4, .2, .1))
plot(cars)  # a scatterplot
```

Note: 

* `cars-plot` is the label of your chunk and of the figure produced by the chunk. You reference it from anywhere in the text with `\@ref(fig:cars-plot)` (of course you can/must replace `cars-plot` with something else)

* `fig.cap="This is your caption"` defines your caption. Always caption your figures and tables.

## How do I create and reference my tables?

To create tables for your PDF document you can use `knitr::kable(my_dataframe)`, where `my_dataframe` is any data frame you might have created and with the values you want to table.

```{r my-table-1}

my_dataframe <- 
  data.frame("col1" = c(1, 2, 3),
             "col2" = c("one", "two", "three"))

knitr::kable(my_dataframe, caption = "This is your table's caption")

```

You can reference the tables you created by (for example, see Table \@ref(tab:my-table-1)) using `\@ref(tab:my-table-1)`.



