---
title: "[SSPS4102 A1 template: Replace with a meaningful title]"
author: "[STUDENT NAME: Replace with your name]"
output:
  bookdown::pdf_document2: default
date: "`r Sys.Date()`"
---

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

```

<!-- Delete from this line until the end of the document before submitting --> 

## How do I create and reference my figures?

See Figure \@ref(fig:cars-plot).

```{r cars-plot, fig.cap="This is your figure's caption"}
par(mar = c(4, 4, .2, .1))
plot(cars)  # a scatterplot
```

Note: 

* `cars-plot` is the label of your chunk and of the figure produced by the chunk. You reference it from anywhere in the text with `\@ref(fig:cars-plot)` (of course you can/must replace `cars-plot` with something else)

* `fig.cap="This is your caption"` defines your caption. Always caption your figures and tables.

## How do I create and reference my tables?

To create tables for your PDF document you can use `knitr::kable(my_dataframe)`, where `my_dataframe` is any data frame you might have created and with the values you want to table.

```{r my-table-1}

my_dataframe <- 
  data.frame("col1" = c(1, 2, 3),
             "col2" = c("one", "two", "three"))

knitr::kable(my_dataframe, caption = "This is your table's caption")

```

You can reference the tables you created by (for example, see Table \@ref(tab:my-table-1)) using `\@ref(tab:my-table-1)`.

### How do I create regression tables?

For regression tables, you have a at least two options. You can simply do this and get Table \@ref(tab:my-table-2).

```{r my-table-2}

fit <- lm(dist ~ speed, data = cars)

knitr::kable(summary(fit)$coef, caption = "These are your coefficients")

```

Or instead, for a more "academic-styled" table like Table \@ref(tab:my-table-3) you can use the [modelsummary](https://vincentarelbundock.github.io/modelsummary/articles/modelsummary.html) package and its many options. Note that the first time you use `modelsummary::modelsummary`, the package will install some required software (without asking).

```{r my-table-3}

fit1 <- lm(dist ~ speed, data = cars)

fit2 <- lm(speed ~ dist, data = cars)

modelsummary::modelsummary(list("OLS 1" = fit1, 
                                "OLS 2" = fit2), 
                           output = "markdown", # This option is important
                           stars = TRUE,
                           title = 'This is my regression.')
```

## How can I include my bibliographic references?

You are more than welcome to style your references manually and to add the full bibliographic reference at the end. Yet if you use a reference management software - something you should definitely consider doing - (e.g. Zotero, EndNote, etc.) and if you really want to master your R Markdown, have a look [here](https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html).  

